---
title: "About"
editor: visual
toc-title: Contents
toc-location: left
toc-depth: 2
code-tools: 
  source: true
  toggle: true
  caption: view source code
code-block-bg: true
code-block-border-left: "#31BAE9"
---
```{r}
#| include: false
library(tidyverse)
library(kableExtra)
```

This analysis evaluates the spatial distribution of power outages in surrounding Houston, Texas caused by winter storms in February 2021 and explores which socioeconomic variables are most correlated with power outages. Code and outputs for the geospatial and statistical analyses can be view from the navigation bar at the top of this page.

:::{.callout-note}
This analysis started as an assignment for a [Spatial Analysis](https://jamesfrew.github.io/EDS_223_spatial_analysis/assignments/2/HW2.html) class I took as part of my [Masters of Environmental Data Science](https://ucsb-meds.github.io/) degree at UC Santa Barbara. Here, the original assignment is expanded upon to include additional socioeconomic variables, a statistical analysis, and interactive leaflet maps.
:::

# Background
In 2021, large portions of the United States experience three major winter storms on February 10-11, February 13-17, and February 15-20. The electrical infrastructure in the state of Texas was particularly effected by these events. In the Houston area, an estimated 1.4 million customers were without power on February 15. Using satellite imagery and census data, this analysis quantifies the magnitude and socioeconomic variability of power outages surrounding Houston. Goals of the analysis include:

- estimate the number of homes without power
- evaluate race, age, and income differences between areas with and without power

# Data
## Satellite imagery

Nighttime light data was obtained from the Visible Infrared Imaging Radiometer Suite (VIIRS), located on the Suomi NPP satellite, which used a low-light sensor to measure light emissions and reflections.

DNB_At_Sensor_Radiance_500m dataset
DNB = Day Night Band

explain file name

Imagery from February 7 and February 16 were used to visualize pre and post storm conditions due to a lack of cloud cover. Imagery from two tiles was needed to cover the area of interest. This results in a total of four files used in the analysis.

- VNP46A1.A2021038.h08v05.001.2021039064328.h5
- VNP46A1.A2021038.h08v06.001.2021039064329.h5
- VNP46A1.A2021047.h08v05.001.2021048091106.h5
- VNP46A1.A2021047.h08v06.001.2021048091105.h5

```{r}
#| echo: false
tibble(component = c("VNP46A1", "A2021038", "h08", "v05", "2021039064328", "h5"),
       meaning = c("name of data product", "year (ie. 2021) and day of year (ie. 038) that the imagery was acquired", "horizontal tile number 8", "vertical tile number 5", "year (2021), day of year (039), and time(06:43:28 UTC) that the file contents were generated", "file format (HDF-EOS5")) %>% 
  kable(col.names = c("file name part", "explanation"),
        caption = "File name convention") %>% 
  kable_paper()
```
 

## Buildings and roads

Building and road data was obtained from [OpenStreetMap](https://www.openstreetmap.org/#map=5/38.007/-95.844). Due to computation constraints, a subset of the road data was provided as a GeoPackage (.gpkg) for the Houston metropolitan area. An SQL query was then used to subset highways and major roads from this GeoPackage.

## Socioeconomic variables

Data pertaining to race, age, and income demographics for Texas census tract was obtained from the U.S. Census Bureau's [American Community Survey](https://www.census.gov/programs-surveys/acs). The data used in this analysis is from 2019 and was cropped to the region of interest. Specific variables evaluated include:

- Race
  - white
  - black
  - native american
  - hispanic / latino
- Age
  - 65 and older
  - 18 and younger
- Income
  - households below poverty level
  - median income


## Limitations
This analysis does not include any verification of power outage locations nor does it incorporate typical variations in nighttime lights. The intent of the statistical analysis is not to use sociodemographic variables as predictors of power outages, but rather provide insight into groups that were disproportionately affected. Further work could incorporate electric utility data to identify vulnerable power grid infrastructure and then recommend equitable asset management and capital improvement plans.


# References

https://blackmarble.gsfc.nasa.gov/tools/OpenHDF5.py

Pebesma E (2022). stars: Spatiotemporal Arrays, Raster and Vector Data Cubes. https://r-spatial.github.io/stars/, https://github.com/r-spatial/stars/.

https://en.wikipedia.org/wiki/2021_Texas_power_crisis

https://earthobservatory.nasa.gov/images/147941/extreme-winter-weather-causes-us-blackouts
