---
title: "Maps"
editor: visual
toc-title: Contents
toc-location: left
toc-depth: 2
code-tools: 
  source: true
  toggle: true
  caption: view source code
code-block-bg: true
code-block-border-left: "#31BAE9"
---

# Population 

```{r}
#| code-fold: true
# leaflet(blackout_buildings_4326) %>%
#   setView(lat = 29.75, lng = -95.5, zoom = 9) %>%
#   addProviderTiles(providers$OpenStreetMap) %>%
#   addPolygons(label = ~paste("Building Type:", type),
#               highlight = highlightOptions(weight = 2, color = "black", bringToFront = TRUE))
```

# pal_pop <- colorBin("Blues", bins = 5, domain = census_tract_data_roi_4326$total_population)
# pal_pop <- colorNumeric("YlGn", domain = census_tract_data_4326$total_population)

```{r}
#| code-fold: true
#pal_blackout <- colorNumeric(c("red"), 1, na.color = "transparent")

#pal_pop <- colorQuantile("Blues", census_tract_data_roi_4326$total_population, n = 5)
# pal_pop <- colorBin("Blues", bins = 5, domain = census_tract_data_roi_4326$total_population)
# 
# leaflet(census_tract_data_roi_4326) %>%
#   addProviderTiles(providers$OpenStreetMap) %>%
#   setView(lat = 29.75, lng = -95.5, zoom = 9) %>%
#   addPolygons(fillColor = ~pal_pop(total_population), fillOpacity = 0.75,
#               color = "gray", weight = 0.75,
#               label = ~paste("Total Population:", format(total_population, big.mark = ",")),
#               highlight = highlightOptions(weight = 3, color = "red", bringToFront = TRUE),
#               group = "group_1") %>%
#   addLegend(pal = pal_pop, values = ~total_population,
#             title = "Total Population", position = "bottomright",
#             group = "group_1") %>%
#   addPolygons(data = blackout_no_highway_4326, fillColor = ~pal_blackout(DNB_At_Sensor_Radiance_500m), fillOpacity = 1, weight = 0,
#               group = "group_2")
  # addLegend(pal = pal_blackout, values = ~blackout_no_highway_4326$DNB_At_Sensor_Radiance_500m,
  #           title = "Blackout Areas", position = "bottomleft",
  #           group = "group_2") %>%
```