---
title: "Statistical Analysis"
editor: visual
toc-title: Contents
toc-location: left
code-tools: 
  source: true
  toggle: true
  caption: view source code
code-block-bg: true
code-block-border-left: "#31BAE9"
---

```{r}
#| warning: false
# load libraries used in analysis
library(tidyverse)
library(patchwork)
library(jtools)
library(kableExtra)
```

```{r}
#| code-overflow: wrap
#| output: false
# read in data generated from geospatial analysis
blackout_census_data <- read_csv("data/census_tract_blackout_data_df.csv")
```

# Linear Regression

Linear regression was used to model the relationship between percent of houses that lost power and selected data from the U.S. Census Bureau's American Community Survey (ACS).

## Race
### Percent white
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population white in census tract"
# linear regression model
model_pct_white <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_white)

#plot
plot_model_pct_white <- ggplot(data = blackout_census_data, aes(x = pct_white, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% white", y = "% of houses that lost power")
plot_model_pct_white
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_white <- summary(model_pct_white)

# extract model coefficients
slope_coef_white <- round(summary_model_pct_white$coefficients["pct_white", "Estimate"], digits = 3)

std_err_white <- round(summary_model_pct_white$coefficients["pct_white", "Std. Error"], digits = 3)

p_value_white <- format(summary_model_pct_white$coefficients["pct_white", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_white <- round(confint(model_pct_white, level = 0.95)[2, 1], digits = 3)

ci_95_upper_white <- round(confint(model_pct_white, level = 0.95)[2, 2], digits = 3)
```

### Percent black
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population black in census tract"
# linear regression model
model_pct_black <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_black)

# plot
plot_model_pct_black <- ggplot(data = blackout_census_data, aes(x = pct_black, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% black", y = "% of houses that lost power")
plot_model_pct_black
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_black <- summary(model_pct_black)

# extract model coefficients
slope_coef_black <- round(summary_model_pct_black$coefficients["pct_black", "Estimate"], digits = 3)

std_err_black <- round(summary_model_pct_black$coefficients["pct_black", "Std. Error"], digits = 3)

p_value_black <- format(summary_model_pct_black$coefficients["pct_black", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_black <- round(confint(model_pct_black, level = 0.95)[2, 1], digits = 3)

ci_95_upper_black <- round(confint(model_pct_black, level = 0.95)[2, 2], digits = 3)
```

### Percent native american
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population native american in census tract"
# linear regression model
model_pct_am_native <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_am_native)

# plot
plot_model_pct_am_native <- ggplot(data = blackout_census_data, aes(x = pct_am_native, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% native american", y = "% of houses that lost power")
plot_model_pct_am_native
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_am_native <- summary(model_pct_am_native)

# extract model coefficients
slope_coef_am_native <- round(summary_model_pct_am_native$coefficients["pct_am_native", "Estimate"], digits = 3)

std_err_am_native <- round(summary_model_pct_am_native$coefficients["pct_am_native", "Std. Error"], digits = 3)

p_value_am_native <- format(summary_model_pct_am_native$coefficients["pct_am_native", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_am_native <- round(confint(model_pct_am_native, level = 0.95)[2, 1], digits = 3)

ci_95_upper_am_native <- round(confint(model_pct_am_native, level = 0.95)[2, 2], digits = 3)
```

### Percent asian
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population asian in census tract"
# linear regression model
model_pct_asian <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_asian)

# plot
plot_model_pct_asian <- ggplot(data = blackout_census_data, aes(x = pct_asian, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% asian", y = "% of houses that lost power")
plot_model_pct_asian
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_asian <- summary(model_pct_asian)

# extract model coefficients
slope_coef_asian <- round(summary_model_pct_asian$coefficients["pct_asian", "Estimate"], digits = 3)

std_err_asian <- round(summary_model_pct_asian$coefficients["pct_asian", "Std. Error"], digits = 3)

p_value_asian <- format(summary_model_pct_asian$coefficients["pct_asian", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_asian <- round(confint(model_pct_asian, level = 0.95)[2, 1], digits = 3)

ci_95_upper_asian <- round(confint(model_pct_asian, level = 0.95)[2, 2], digits = 3)
```

### Percent hispanic / latino
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population hispanic / latino in census tract"
# linear regression model
model_pct_hispanic_latino <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_hispanic_latino)

# plot
plot_model_pct_hispanic_latino <- ggplot(data = blackout_census_data, aes(x = pct_hispanic_latino, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% hispanic / latino", y = "% of houses that lost power")
plot_model_pct_hispanic_latino
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_hispanic_latino <- summary(model_pct_hispanic_latino)

# extract model coefficients
slope_coef_hispanic_latino <- round(summary_model_pct_hispanic_latino$coefficients["pct_hispanic_latino", "Estimate"], digits = 3)

std_err_hispanic_latino <- round(summary_model_pct_hispanic_latino$coefficients["pct_hispanic_latino", "Std. Error"], digits = 3)

p_value_hispanic_latino <- format(summary_model_pct_hispanic_latino$coefficients["pct_hispanic_latino", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_hispanic_latino <- round(confint(model_pct_hispanic_latino, level = 0.95)[2, 1], digits = 3)

ci_95_upper_hispanic_latino <- round(confint(model_pct_hispanic_latino, level = 0.95)[2, 2], digits = 3)
```

### Race summary plots
```{r fig.width=10,fig.height=5}
#| warning: false
#| code-overflow: wrap
#| column: screen-right
model_plots_race <- (plot_model_pct_white | plot_model_pct_black | plot_model_pct_am_native) / 
  (plot_model_pct_asian | plot_model_pct_hispanic_latino | plot_spacer()) +
  plot_annotation(title = "Lineral Regression Models of Power Outages vs. Race")

model_plots_race
```

## Age
### Percent 65 and older
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population age 65 and over in census tract"
# linear regression model
model_pct_65_and_over <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_65_and_over)

# plot
plot_model_pct_65_and_over <- ggplot(data = blackout_census_data, aes(x = pct_65_and_over, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% 65 and over", y = "% of houses that lost power")
plot_model_pct_65_and_over
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_65_and_over <- summary(model_pct_65_and_over)

# extract model coefficients
slope_coef_pct_65_and_over <- round(summary_model_pct_65_and_over$coefficients["pct_65_and_over", "Estimate"], digits = 3)

std_err_pct_65_and_over <- round(summary_model_pct_65_and_over$coefficients["pct_65_and_over", "Std. Error"], digits = 3)

p_value_pct_65_and_over <- format(summary_model_pct_65_and_over$coefficients["pct_65_and_over", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_pct_65_and_over <- round(confint(model_pct_65_and_over, level = 0.95)[2, 1], digits = 3)

ci_95_upper_pct_65_and_over <- round(confint(model_pct_65_and_over, level = 0.95)[2, 2], digits = 3)
```

### Percent children under 18
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population age 18 and younger in census tract"
# linear regression model
model_pct_children_under_18 <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_children_under_18)

# plot
plot_model_pct_children_under_18 <- ggplot(data = blackout_census_data, aes(x = pct_children_under_18, y = pct_houses_that_lost_power)) +
  geom_point() +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% children under 18", y = "% of houses that lost power")
plot_model_pct_children_under_18
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_children_under_18 <- summary(model_pct_children_under_18)

# extract model coefficients
slope_coef_children_under_18 <- round(summary_model_pct_children_under_18$coefficients["pct_children_under_18", "Estimate"], digits = 3)

std_err_children_under_18 <- round(summary_model_pct_children_under_18$coefficients["pct_children_under_18", "Std. Error"], digits = 3)

p_value_children_under_18 <- format(summary_model_pct_children_under_18$coefficients["pct_children_under_18", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_children_under_18 <- round(confint(model_pct_children_under_18, level = 0.95)[2, 1], digits = 3)

ci_95_upper_children_under_18 <- round(confint(model_pct_children_under_18, level = 0.95)[2, 2], digits = 3)
```

### Age summary plots
```{r fig.width=10,fig.height=3}
#| warning: false
#| code-overflow: wrap
#| column: screen-right

model_plots_age <- (plot_model_pct_65_and_over | plot_model_pct_children_under_18) +
  plot_annotation(title = "Lineral Regression Models of Power Outages vs. Age")

model_plots_age
```

## Income
### Percent households below poverty
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. percent population of household below the poverty level in census tract"
# linear regression model
model_pct_households_below_poverty <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ pct_households_below_poverty)

# plot
plot_model_pct_households_below_poverty <- ggplot(data = blackout_census_data, aes(x = pct_households_below_poverty, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "% households below poverty", y = "% of houses that lost power")
plot_model_pct_households_below_poverty
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_pct_households_below_poverty <- summary(model_pct_households_below_poverty)

# extract model coefficients
slope_coef_households_below_poverty <- round(summary_model_pct_households_below_poverty$coefficients["pct_households_below_poverty", "Estimate"], digits = 3)

std_err_households_below_poverty <- round(summary_model_pct_households_below_poverty$coefficients["pct_households_below_poverty", "Std. Error"], digits = 3)

p_value_households_below_poverty <- format(summary_model_pct_households_below_poverty$coefficients["pct_households_below_poverty", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_households_below_poverty <- round(confint(model_pct_households_below_poverty, level = 0.95)[2, 1], digits = 3)

ci_95_upper_households_below_poverty <- round(confint(model_pct_households_below_poverty, level = 0.95)[2, 2], digits = 3)
```

### Median income
```{r}
#| warning: false
#| code-overflow: wrap
#| code-fold: false
#| column: margin
#| fig-cap: "Percent of houses that lost power vs. median income in census tract"
# linear regression model
model_median_income <- lm(data = blackout_census_data, pct_houses_that_lost_power ~ median_income)

# plot
plot_model_median_income <- ggplot(data = blackout_census_data, aes(x = median_income, y = pct_houses_that_lost_power)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm, formula = y~x, se = FALSE) +
  theme_classic() +
  labs(x = "median income", y = "% of houses that lost power")
plot_model_median_income
```

```{r}
#| code-overflow: wrap
#| code-fold: true
# model summary
summary_model_median_income <- summary(model_median_income)

# extract model coefficients
slope_coef_median_income <- round(summary_model_median_income$coefficients["median_income", "Estimate"], digits = 3)

std_err_median_income <- round(summary_model_median_income$coefficients["median_income", "Std. Error"], digits = 3)

p_value_median_income <- format(summary_model_median_income$coefficients["median_income", "Pr(>|t|)"], scientific = FALSE, digits = 2)

# confidence intervals
ci_95_lower_median_income <- round(confint(model_median_income, level = 0.95)[2, 1], digits = 3)

ci_95_upper_median_income <- round(confint(model_median_income, level = 0.95)[2, 2], digits = 3)
```

### Income summary plots
```{r fig.width=10,fig.height=3}
#| warning: false
#| code-overflow: wrap
#| column: screen-right

model_plots_income <- (plot_model_pct_households_below_poverty | plot_model_median_income) +
  plot_annotation(title = "Lineral Regression Models of Power Outages vs. Income")

model_plots_income
```


# Results
```{r}
#| code-overflow: wrap
#| message: false
race_model_summary <- plot_summs(model_pct_white, model_pct_black, model_pct_am_native, model_pct_asian, model_pct_hispanic_latino,
           inner_ci_level = 0.9,
           model.names = c("white", "black", "native american", "asian", "hispanic/latino"))
race_model_summary
```

```{r}
#| message: false
#| #| code-overflow: wrap
age_model_summary <-plot_summs(model_pct_65_and_over, model_pct_children_under_18,
           inner_ci_level = 0.95,
           model.names = c("over 65", "under 18"))
age_model_summary
```

```{r}
#| message: false
#| code-overflow: wrap
income_model_summary <-plot_summs(model_pct_households_below_poverty, model_median_income,
           inner_ci_level = 0.95,
           model.names = c("poverty", "median income"))
income_model_summary
```

```{r}
#| code-overflow: wrap
#| code-fold: true
#| column: page-right
model_summary_df <- data.frame(model_name = c("white", "black", "native american", "asian", "hispanic / latino", "65 and over", "18 and younger", "poverty", "median income"),
                               slope_coef = c(slope_coef_white, slope_coef_black, slope_coef_am_native, slope_coef_asian, slope_coef_hispanic_latino, slope_coef_pct_65_and_over, slope_coef_children_under_18, slope_coef_households_below_poverty, slope_coef_median_income),
                               std_err = c(std_err_white, std_err_black, std_err_am_native, std_err_asian, std_err_hispanic_latino, std_err_pct_65_and_over, std_err_children_under_18, std_err_households_below_poverty, std_err_median_income),
                               p_value = c(p_value_white, p_value_black, p_value_am_native, p_value_asian, p_value_hispanic_latino, p_value_pct_65_and_over, p_value_children_under_18, p_value_households_below_poverty, p_value_median_income),
                               ci_95_lower = c(ci_95_lower_white, ci_95_lower_black, ci_95_lower_am_native, ci_95_lower_asian, ci_95_lower_hispanic_latino, ci_95_lower_pct_65_and_over, ci_95_lower_children_under_18, ci_95_lower_households_below_poverty, ci_95_lower_median_income),
                               ci_95_upper = c(ci_95_upper_white, ci_95_upper_black, ci_95_upper_am_native, ci_95_upper_asian, ci_95_upper_hispanic_latino, ci_95_upper_pct_65_and_over, ci_95_upper_children_under_18, ci_95_upper_households_below_poverty, ci_95_upper_median_income)) %>% 
  mutate(significance = case_when(
    p_value < 0.05 ~ "yes",
    p_value >= 0.05 ~ "no"))

model_summary_table <- model_summary_df %>% 
  kable(col.names = c("model name", "slope coefficent", "standard error", "p-value", "95% confidence\ninterval (lower)", "95% confidence\nintervaln(upper)", "significant\nat 5% level"),
        caption = "Summary of Model Results") %>% 
  kable_paper()
model_summary_table
```




# Discussion

Note: the purpose of this evaluation was not to predict houses that could be more likely to lose power in the future. This analysis attempts to identify disproportional vulnerabilities of the effected community.
